### 2017-07-02 abstract.tex

  To address this issue, we present an automated verification
  methodology based on a novel program logic that enables
  compositional reasoning about the behavior of concurrently executing
  weakly-isolated transactions. Notably, our development is parametric
  over a transaction's specific isolation semantics, allowing it to be
  applicable over a range of concurrency control mechanisms.  Given a
  a collection of concurrently executing transactions, our analysis
  verifies whether assertion ${\mathcal{A}}$ associated with
  transaction $T$ in this collection, holds under a weak isolation
  level ${\mathcal{I}}$.  Case studies and experiments on real-world
  applications demonstrate the utility of our approach, and provide
  strong evidence that verification of weakly-isolated transactions
  can be placed on the same formal footing as their strongly-isolated
  serializable counterparts.

>>>>

  To address this issue, we present a novel program logic that enables
  compositional reasoning about the behavior of concurrently executing
  weakly-isolated transactions.  Recognizing that the proof burden
  necessary to use this logic may dissuade application developers, we
  also describe a fully automated verification tool based on this
  foundation that determines the safety of high-level consistency
  assertions associated with such transactions.  The key to effective
  inference is the observation that weakly-isolated transactions can
  be viewed as functional (monadic) computations over database state,
  allowing us to treat their operations as state transformers over the
  database.  This interpretation enables automated verification using
  off-the-shelf SMT solvers.

  Notably, our development is parametric over a transaction's specific
  isolation semantics, allowing it to be applicable over a range of
  concurrency control mechanisms.  Case studies and experiments on
  real-world applications demonstrate the utility of our approach, and
  provide strong evidence that automated verification of
  weakly-isolated transactions can be placed on the same formal
  footing as their strongly-isolated serializable counterparts.

### 2017-07-02 introduction.tex

The guarantees provided by serializability do not come for free,
however - pessimistic concurrency control methods that enforce
serializability, for example, require databases to provide expensive
techniques such as two-phase locking that incur overhead to deal with
deadlocks, rollbacks, and re-execution~\cite{twopl,ullmanbook}.
Similar criticisms apply to optimistic multi-version concurrency
control methods that must deal with timestamp and version
management~\cite{BG81}.  Regardless of the particular concurrency
control method employed, enforcing serializability in a replicated
data store additionally requires global coordination among
geo-distributed replicas, which makes the store \emph{unavailable} in
the presence of network
partitions~\cite{cap,sernotavlbl,bailishat,bernsigmod13}.

The tension between serializable transactions, which are easy to
reason about but difficult to implement, and more pragmatic variants
that are driven by performance and availability considerations, has
motivated the development of weaker forms of transaction isolation,
beginning as early as 1976~\cite{gray1976}. The ANSI SQL 92 standard
defines three such weak isolation levels which are now implemented in
many relational and NoSQL databases. Not surprisingly, weakly-isolated
transactions have been found to significantly outperform serializable
transactions on benchmark suites, both on single-node databases and
multi-node replicated stores~\cite{dbtuningbook,bailishat,bailisvldb},
leading to their overwhelming adoption. A 2013
study~\cite{bailishotos} of 18 popular ACID and ``NewSQL'' databases
found that only three of them offer serializability by default, and
half, including Oracle 11g, do not offer it at all.  A 2015
study~\cite{bailisferal} of a large corpus of database applications
finds no evidence that applications manifestly change the default
isolation level offered by the database. Taken together, these studies
make clear that weakly-isolated transactions are quite prevalent in
practice, whereas serializable transactions are often eschewed.

Unfortunately, weak isolation admits behaviours that are difficult to
comprehend~\cite{berenson}. To quantify weak isolation anomalies,
Fekete \emph{et al.}~\cite{feketevldb09} devised and experimented
with a microbenchmark suite that executes transactions under a
weakly-isolated \emph{read committed} isolation level - the default
level for 8 of the 18 databases studied in~\cite{bailishotos}, and
found that 25 out of every 1000 rows in the database violate at least
one integrity constraint. Bailis \emph{et al.}~\cite{bailisferal} rely
on Rails' \emph{uniqueness validation} to maintain uniqueness of
records while serving Linkbench's~\cite{linkbench} insertion workload
(6400 records distributed over 1000 keys; 64 concurrent clients), and
report discovering more than 10 duplicate records.  Rails relies on
database transactions to validate uniqueness during insertions, which
is sensible if transactions are serializable, but incorrect under the
\emph{read committed} isolation level used in the experiments. The
same study has found that 13\% of all invariants among 67 open source
Ruby-on-Rails applications are at risk of being violated due to weak
isolation. Indeed, incidents of safety violations due to executing
applications in a weakly-isolated environment have been reported on
web services in production~\cite{starbucksbug, scimedbug}, including
in safety-critical applications such as bitcoin
exchanges~\cite{poloniexbug, bitcoinbug}. While enforcing
serializability for all transactions would be sufficient to avoid
these errors and anomalies, it would likely be an overly conservative
strategy; indeed, 75\% of the invariants studied in~\cite{bailisferal}
were shown to be preserved under some form of weak isolation.  When to
use weak isolation, and in what form, is therefore a prominent
question facing all database programmers.

A major problem with weak isolation is that its semantics in the
context of user programs is not well-understood. The original
proposal~\cite{gray1976} defines multiple ``degrees'' of weak
isolation in terms of implementation details such as the nature and
duration of locks held in each case. The ANSI SQL 92 standard defines
four levels of isolation (including serializability) in terms of
various undesirable \emph{phenomena} (\eg \emph{dirty reads}) each is
required to prevent. While this is an improvement, it requires
programmers to be prescient about the possible ways various
undesirable phenomena might manifest in their applications, and in
each case determine if the phenomenon can be allowed without violating
application invariants. This is understandably hard, especially in the
absence of any formal underpinning to define weak isolation semantics.
Adya~\cite{adyaphd} presents the first formal definitions of some
well-known isolation levels in the context of a sequentially
consistent (SC) database.  However, there has been little progress
relating Adya's system model to a formal operational semantics or a
proof system that can facilitate rigorous correctness arguments.
Consequently, reasoning about weak isolation remains an error prone
endeavour, with major database vendors~\cite{postgresiso, mysqliso,
oracleiso} continuing to document their isolation levels primarily in
terms of the undesirable phenomena a particular isolation level may
induce, placing the burden on the programmer to determine application
correctness.

%%%SJ: tried to tighten this paragraph; the original now commented seemed overly
%%complicated
Recent results on reasoning about application invariants in the
presence of weak consistency~\cite{burckhardt14, redblueosdi,
redblueatc, ecinec, gotsmanpopl16} address broadly related concerns.
Weak consistency is a phenomenon that manifests on replicated data
stores, where atomic operations are concurrently executed against
different replicas, resulting in an execution order inconsistent with
%% We already say "inconsistent with.. ". 
any sequential order. In contrast, weak isolation is a property of
concurrent transactions interfering with one another resulting in an
execution order that is not serializable. Unlike weak consistency,
weak isolation can manifest even in an unreplicated setting, as
evident from the support for weakly-isolated transactions on
conventional (unreplicated) databases as mentioned above. In the
presence of replication, however, the interaction between weak
isolation and weak consistency can be subtle and non-trivial.
Understanding weak isolation in these varied contexts thus requires
new insights and substantial generalization of existing techniques.

@@

In this paper, we propose a program logic for weakly-isolated
transactions that realizes this goal.  In particular, we develop a set
of syntax-directed compositional proof rules that allow developers to
build correctness proofs for transactional programs in the presence of
a weakly-isolated concurrency control mechanism.  A key novelty of our
approach is that it is parametric over the \emph{isolation semantics}
of transactions in the program, as well as the \emph{consistency
semantics} of the underlying store. In concrete terms, this means
that, unlike recent work focused on reasoning about programs under
weak consistency~\cite{gotsmanpopl16, redblueatc, ecinec}, our system
model does not assume a minimum or predefined set of consistency or
isolation levels. Instead, our operational semantics admits
declarative specifications of transaction isolation and store
consistency, and generates executions that are guaranteed to satisfy
these specifications. The result is a flexible system model that is
general enough to incorporate the semantics of a range of isolation
levels on a variety of stores (\eg~the sequentially consistent store
of~\cite{adyaphd}, or the causally consistent store
of~\cite{gotsmanpopl16}, etc).  Our key technical contribution is thus
a set of proof rules that demonstrate that a program with a given
selection of isolation levels for its transactions preserves its
invariants when executed on a store equipped with appropriate consistency
guarantees.  The paper makes the following contributions:
\vspace*{-6pt}
\begin{enumerate}
  \item We develop a semantics for a core language equipped with
    weakly-isolated transactions, demonstrating that a general
    parametric isolation semantics can be expressed as
    well-formedness constraints on a program's execution.
  \item We present a compositional proof system for this language
    capable of relating high-level application invariants to the structure
    of traces induced by the operational semantics, in which transactions
    are associated with specific weak isolation levels.
  \item We define a \emph{maximum visibility principle} to reconcile
    store consistency guarantees with transaction isolation
    requirements, and use it to instantiate our operational model and
    the proof system for various kinds of (weakly-consistent) data
    stores.
  \item Case studies modeled after real-world scenarios demonstrate
    the applicability and utility of our proof methodology.
\end{enumerate}
\noindent Our results provide the first (to the best of our knowledge)
mechanism that precisely and uniformly relates high-level program
invariants to low-level weak isolation and weak consistency
guarantees, thereby allowing weakly-isolated transactions to enjoy the
same rigorous reasoning capabilities as their strongly-isolated
(serializable) counterparts.

The remainder of the paper is organized as follows. The next section
provides motivation and background on serializable and weakly-isolated
transactions. \S\ref{sec:opsem} presents an operational semantics for
a core language that supports weakly-isolated transactions,
parameterized over different isolation notions. \S\ref{sec:reasoning}
formalizes the proof system that we use to reason about program
invariants, and establishes the soundness of these rules with respect
to the semantics. \S\ref{sec:store-consistency} generalizes the
framework to integrate support for weakly-consistent data stores. We
describe the impact of our reasoning framework in the context of
several real-world case studies in \S\ref{sec:case-studies}.  Related
work and conclusions are given in \S\ref{sec:relatedwork}.

>>>>

The guarantees provided by serializability do not come for free,
however - pessimistic concurrency control methods require databases to
use expensive mechanisms such as two-phase locking that incur overhead
to deal with deadlocks, rollbacks, and
re-execution~\cite{twopl,ullmanbook}.  Similar criticisms apply to
optimistic multi-version concurrency control methods that must deal
with timestamp and version management~\cite{BG81}.  These issues are
further exacerbated when the database is replicated, requiring
additional coordination
mechanisms~\cite{cap,sernotavlbl,bailishat,bernsigmod13}.

The tension between serializable transactions, which are easy to
reason about but difficult to implement efficiently, and more
pragmatic variants that are driven by performance and availability
considerations, has motivated the development of weaker forms of
transaction isolation, beginning as early as 1976~\cite{gray1976}. The
ANSI SQL 92 standard defines three such weak isolation levels which
are now implemented in many relational and NoSQL databases. Not
surprisingly, weakly-isolated transactions have been found to
significantly outperform serializable transactions on benchmark
suites, both on single-node databases and multi-node replicated
stores~\cite{dbtuningbook,bailishat,bailisvldb}, leading to their
overwhelming adoption. A 2013 study~\cite{bailishotos} of 18 popular
ACID and ``NewSQL'' databases found that only three of them offer
serializability by default, and half, including Oracle 11g, do not
offer it at all.  A 2015 study~\cite{bailisferal} of a large corpus of
database applications finds no evidence that applications manifestly
change the default isolation level offered by the database. Taken
together, these studies make clear that weakly-isolated transactions
are quite prevalent in practice, whereas serializable transactions are
often eschewed.

Unfortunately, weak isolation admits behaviors that are difficult to
comprehend~\cite{berenson}. To quantify weak isolation anomalies,
Fekete \emph{et al.}~\cite{feketevldb09} devised and experimented with
a microbenchmark suite that executes transactions under a
weakly-isolated \emph{read committed} isolation level - the default
level for 8 of the 18 databases studied in~\cite{bailishotos}, and
found that 25 out of every 1000 rows in the database violate at least
one integrity constraint. Bailis \emph{et al.}~\cite{bailisferal} rely
on Rails' \emph{uniqueness validation} to maintain uniqueness of
records while serving Linkbench's~\cite{linkbench} insertion workload
(6400 records distributed over 1000 keys; 64 concurrent clients), and
report discovering more than 10 duplicate records.  Rails relies on
database transactions to validate uniqueness during insertions, which
is sensible if transactions are serializable, but incorrect under the
\emph{read committed} isolation level used in the experiments. The
same study has found that 13\% of all invariants among 67 open source
Ruby-on-Rails applications are at risk of being violated due to weak
isolation. Indeed, incidents of safety violations due to executing
applications in a weakly-isolated environment have been reported on
web services in production~\cite{starbucksbug, scimedbug}, including
in safety-critical applications such as bitcoin
exchanges~\cite{poloniexbug, bitcoinbug}. While enforcing
serializability for all transactions would be sufficient to avoid
these errors and anomalies, it would likely be an overly conservative
strategy; indeed, 75\% of the invariants studied in~\cite{bailisferal}
were shown to be preserved under some form of weak isolation.  When to
use weak isolation, and in what form, is therefore a prominent
question facing all database programmers.\footnote{This position has
  been echoed by leading database researchers who lament the lack of
  better understanding of this problem.  See: {\tt
    http://www.bailis.org/blog/understanding-weak-isolation-is-a-serious-problem}.}

A major problem with weak isolation is that its semantics in the
context of user programs is not well-understood. The original
proposal~\cite{gray1976} defines multiple ``degrees'' of weak
isolation in terms of implementation details such as the nature and
duration of locks held in each case. The ANSI SQL 92 standard defines
four levels of isolation (including serializability) in terms of
various undesirable \emph{phenomena} (\eg \emph{dirty reads}) each is
required to prevent. While this is an improvement, it requires
programmers to be prescient about the possible ways various
undesirable phenomena might manifest in their applications, and in
each case determine if the phenomenon can be allowed without violating
application invariants. This is understandably hard, especially in the
absence of any formal underpinning to define weak isolation semantics.
Adya~\cite{adyaphd} presents the first formal definitions of some
well-known isolation levels in the context of a sequentially
consistent (SC) database.  However, there has been little progress
relating Adya's system model to a formal operational semantics or a
proof system that can facilitate rigorous correctness arguments.
Consequently, reasoning about weak isolation remains an error prone
endeavor, with major database vendors~\cite{postgresiso, mysqliso,
oracleiso} continuing to document their isolation levels primarily in
terms of the undesirable phenomena a particular isolation level may
induce, placing the burden on the programmer to determine application
correctness.

Recent results on reasoning about application invariants in the
presence of weak consistency~\cite{burckhardt14, redblueosdi,
redblueatc, ecinec, gotsmanpopl16} address broadly related concerns.
Weak consistency is a phenomenon that manifests on replicated data
stores, where atomic operations are concurrently executed against
different replicas, resulting in an execution order inconsistent with
any sequential order. In contrast, weak isolation is a property of
concurrent transactions interfering with one another resulting in an
execution order that is not serializable. Unlike weak consistency,
weak isolation can manifest even in an unreplicated setting, as
evident from the support for weakly-isolated transactions on
conventional (unreplicated) databases as mentioned above.

%% In the
%% presence of replication, however, the interaction between weak
%% isolation and weak consistency can be subtle and non-trivial.
%% Understanding weak isolation in these varied contexts thus requires
%% new insights and substantial generalization of existing techniques.

@@

In this paper, we propose a program logic for weakly-isolated
transactions along with automated verification support that realizes
this goal.  We develop a set of syntax-directed compositional proof
rules that allow developers to build correctness proofs for
transactional programs in the presence of a weakly-isolated
concurrency control mechanism.  Moreover, we also present an inference
algorithm based on a state transformer semantics that allows us to
automatically verify the weakest isolation level for a transaction
that ensures its consistency requirements are maintained.  The key to
inference is our formulation of database state (represented as sets of
tuples) as a monad, in which database computations are interpreted as
transformers over these sets.  This interpretation leads to an
encoding database operations amenable for verification by SMT solvers.

We have realized these ideas as an embedded DSL within OCaml, with
support for common SQL operations (updates, selects, inserts, deletes,
joins, etc.).  Experimental results on real-world benchmarks
demonstrate the feasibility of automated verification of weakly
isolated transactions, an important advance in improving the safety of
realistic database computations. The paper makes the following
contributions:
\vspace*{-6pt}
\begin{enumerate}
\item An operational semantics for an embedded DSL in OCaml that
  expresses SQL-based relational database operations as monadic
  computations on database state represented abstractly as sets of
  tuples.
 \item A compositional proof system for this language capable of
   relating high-level application invariants to transformations on
   database state, parameterized by a weak isolation semantics that
   selectively exposes the visibility of these operations to other
   transactions.
 \item An inference algorithm capable of discovering the
   weakest isolation level that is sound with respect to a
   transaction's high-level consistency requirements.  The algorithm
   exploits our monadic formulation, treating database computations as
   predicate state transformers, that leads to a natural encoding
   amenable for automated verification.
 \item A realistic implementation along with results on real-world
   database benchmarks that justify our approach and the utility of
   our proof methodology.
\end{enumerate}
\noindent Our results provide the first (to the best of our knowledge)
automated mechanism to verify the safety of high-level consistency
conditions for weakly-isolated transactions, thereby allowing such
transactions to enjoy the same rigorous reasoning capabilities as
their strongly-isolated (serializable) counterparts.

The remainder of the paper is organized as follows. The next section
provides motivation and background on serializable and weakly-isolated
transactions. \S\ref{sec:opsem} presents an operational semantics for
a core language that supports weakly-isolated transactions,
parameterized over different isolation notions. \S\ref{sec:reasoning}
formalizes the proof system that we use to reason about program
invariants, and establishes the soundness of these rules with respect
to the semantics. \S\ref{sec:inference} describes the inference
algorithm, and the state transformer encoding.  We describe our
implementation, benchmark results, and case studies in
\S\ref{sec:case-studies}.  Related work and conclusions are given in
\S\ref{sec:relatedwork}.

### 2017-07-02 motivation.tex

We have developed and implemented our ideas in an embedded DSL in
OCaml that supports a \C{DB} monad to define and compose database
computations given in terms of SQL operations.  Besides the usual
\C{bind} and \C{return} combinators, the monad offers an
\C{atomically} combinator that executes a database computation as an
atomic transaction and returns the result.

@@

Fig.~\ref{fig:new_order_code} shows a simplified version of the TPC-C
\C{new\_order} transaction written in this language.  TPC-C is a
well-known Online Transaction Processing (OLTP) benchmark that models
an order-processing system for a wholesale parts supply business. The
business logic is captured in 5 database transactions that operate on
9 tables; \C{new\_order} is one such transaction that operates on
\C{District}, \C{Order}, \C{New\_order}, \C{Stock}, and
\C{Order\_line} tables. The transaction acts on the behalf of a
customer, whose id is \C{c\_id}, to place a new order for the given
set of items (\C{item\_reqs}), to be served by a warehouse under the
district identified by \C{d\_id}.  Fig.~\ef{fig:scheme} illustrates
the relationship among these different tables.

@@

Because the aforementioned reasoning is reasonably simple to perform
manually, verfiying the soundess of TPC-C's consistency conditions
would appear to be feasible.  Serializability aids the tractability of
verification by preventing any interference among concurrently
executing transactions while the \C{new\_order} transaction executes.
Under weak isolation\footnote{Weak isolation doesn't violate atomicity
  as long as the witnessed effects are those of committed
  transactions}, however, interferences of various kinds are
permitted.  Although the verification problem for weakly isolated
transactions would appear to be superficially similar to the
verification of (racy) concurrent programs~\cite{concurrentGC...},
weak isolation introduces new challenges arising from the use of
transactions, and new opportunities arising from the fact that the
store abstraction used by transactions is a relational database, not
low-level memory.

@@

This observation leads us away from thinking of database transactions
as concurrent imperative programs.  Instead, we see value in viewing
them as essentially functional computations that manage database state
monadically~\cite{statemonad}. We find it useful to reason about
statements that mutate the database state, not in terms of a pre- and
post-condition pair, but in terms of a state transformer that relates
the pre- and post-states of a statement. This state transformer
semantics can be defined algorithmically, just like predicate
transformer semantics (e.g., strongest post-condition).  Here, a state
transformer interprets a SQL statement in the set domain, taking
advantage of the fact that a a database is essentially a set of
tuples, and a SQL statement is a transformer over these sets.  The
benefit of this approach is that low-level loops can now be
substituted with higher-order combinators that automatically lift the
state transformer of its higher-order argument (i.e., the loop body)
to the state transformer of the combined expression (i.e., the loop).
Thus the semantics of a \C{foreach} loop, for instance, can be
captured as a state transformer, where the state is a set, and the
transformation defines a bind operation. We illustrate this intuition
on a simple example.

>>>>

We present our ideas in the context of an embedded DSL in OCaml that
we have developed which supports a \C{DB} monad to define and compose
database computations given in terms of SQL operations.  Besides the
usual \C{bind} and \C{return} operators, the monad offers an
\C{atomically} combinator that executes a database computation as an
atomic transaction and returns the result.

@@

Fig.~\ref{fig:new_order_code} shows a simplified version of the TPC-C
\C{new\_order} transaction written in this language.  TPC-C is a
well-known Online Transaction Processing (OLTP) benchmark that models
an order-processing system for a wholesale parts supply business. The
business logic is captured in 5 database transactions that operate on
9 tables; \C{new\_order} is one such transaction that uses
\C{District}, \C{Order}, \C{New\_order}, \C{Stock}, and
\C{Order\_line} tables. The transaction acts on the behalf of a
customer, whose id is \C{c\_id}, to place a new order for a given
set of items (\C{item\_reqs}), to be served by a warehouse under the
district identified by \C{d\_id}.  Fig.~\ref{fig:scheme} illustrates
the relationship among these different tables.

@@

Because the aforementioned reasoning is reasonably simple to perform
manually, verfiying the soundess of TPC-C's consistency conditions
would appear to be feasible.  Serializability aids the tractability of
verification by preventing any interference among concurrently
executing transactions while the \C{new\_order} transaction executes.
Under weak isolation\footnote{Weak isolation does not violate
  atomicity as long as the witnessed effects are those of committed
  transactions}, however, interferences of various kinds are
permitted.  Although the verification problem for weakly isolated
transactions would appear to be superficially similar to the
verification of (racy) concurrent programs (e.g., garbage
collectors~\cite{JLP+14,GHE15,HPQ+15}), weak isolation introduces new
challenges arising from the use of transactions, and new opportunities
arising from the fact that the store abstraction used by transactions
is a relational database, not low-level memory.

@@

This observation leads us away from thinking of database transactions
as concurrent imperative programs.  Instead, we see value in viewing
them as essentially functional computations that manage database state
monadically. We find it useful to reason about
statements that mutate the database state, not in terms of a pre- and
post-condition pair, but in terms of a state transformer that relates
the pre- and post-states of a statement. This state transformer
semantics can be defined algorithmically, just like predicate
transformer semantics (e.g., strongest post-condition).  Here, a state
transformer interprets a SQL statement in the set domain, taking
advantage of the fact that a a database is essentially a set of
tuples, and a SQL statement is a transformer over these sets.  The
benefit of this approach is that low-level loops can now be
substituted with higher-order combinators that automatically lift the
state transformer of its higher-order argument (i.e., the loop body)
to the state transformer of the combined expression (i.e., the loop).
Thus the semantics of a \C{foreach} loop, for instance, can be
captured as a state transformer, where the state is a set, and the
transformation defines a bind operation. We illustrate this intuition
on a simple example.
